Co-methylation

### 2014-12-23 Rmarkdown/comethylation.Rmd
Rough draft of co-methylation chapter complete-ish.

\section{Correlations of $\beta$-values}\label{subsec:beta_cor}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{../figures/feature_status.pdf}
\caption{Schematic illustrating the 'feature status' of four pairs of methylation loci. The circles represent loci and the grey bars a genomic feature, such as CpG islands. Each pair's 'feature status' is determined by whether both loci are outside of the feature (outside), one locus is inside and one outside of the feature (spanning), or both inside the feature (inside). Note that the rightmost pair is 'inside' even though the two loci that comprise the pair are in different elements of the genomic feature.}
\label{fig:feature_status}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/ADS_beta_scatterplots.pdf}
\caption{Kernel density smoothed scatterplots for pairs of CpG $\beta$-values, $(\beta_{1}, \beta_{2})$, from the ADS sample. $\beta$-values are computed using strand-collapsed counts. $n$ is the number of pairs in each scatterplot.  The Spearman correlation of each scatterplot is also reported. Individual plots created using the \texttt{smoothScatter} function in R.}
\label{fig:ADS_beta_scatterplots}
\end{figure}

\subsection{Results}

@@

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show that the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand. Since the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_5}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\subsubsection{$NIL = 0$}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\subsubsection{$NIL \geq 0$}\label{sec:nil_geq_0_beta_cors}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

The $150-200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands. __TODO: Discuss more__

\subsection{Summary}

The $NIL = 0$ results measure the 'first-order' or 'one-step' correlations of $\beta$-values whereas the $NIL \geq 0$ measure a mixture of different orders of correlations of $\beta$-values. For this reason I find the $NIL = 0$ plots are easier to interpret, however, they do not tell the whole story.

By comparing the $NIL = 0$ and $NIL \geq 0$ results we see that the strength of the correlation depends on not just the distance between the two CpGs but the number of intervening CpGs. This indicates that these correlations are more than 'first-order'. Moreover, the differences in correlations between CpG islands and non-islands indicates that the 'order' of these correlations itself varies across the genome. All this means that the vector of $\beta$-values for a single sample forms a complicated and highly non-stationary process with variable-order correlations.

This is valuable information in trying to account for correlated $\beta$-values in downstream analyses. However, it is based on multiple levels of aggregation, such as aggregation of cells to form $\beta$-values and such aggregations may hide the source of these correlated $\beta$-values. To better understand what leads to correlated $\beta$-values, I investigated within-fragment co-methylation.

\section{Within-fragment co-methylation}\label{subsec:wf_cometh}

@@

Here I describe a simple method based on analysing methylation patterns at 2-tuples[^higher_order_m-tuples]. For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type. This method is implemented in the `cometh` function that is part of the `MethylationTuples` software.

[^higher_order_m-tuples]: Extensions to m-tuples with $\text{m} > 2$ are discussed in __SECTION__.

@@

\subsubsection{Choice of dependence measure}\label{sec:choice_of_dependence_measure}

@@

There are several alternative measures of association for a $2 \times 2$ table of binary variables such as __EXAMPLES__

__TODO: Comparison of LORs and Pearson correlations of Landan__

- Plot histogram of LOR and histogram of r

- Look for tables with LOR > 0 and r = 0 (and vice versa?)

__TODO: Discussion of Kraemer's work, in particular, the fact that non-zero LOR are not informative for the strength of the correlation__

__TODO: \cite{Bonett:2007bd} is a reference that Pearson correlation of 2x2 is phi coefficient__

\subsection{Results}

$LOR$s are only computed for 2-tuples with sufficient sequencing coverage (here, $5 \times$ for stranded 2-tuples and $10 \times$ for unstranded 2-tuples). For any given $IPD$ there are hundreds of thousands of CpG pairs, and so there is a distribution of co-methylation for any given $IPD$. We could visualise the distribution for a given $IPD$ using classical visualisation techniques such as histograms, kernel density estimates and boxplots. These summaries may be stratified by one or more annotations. However, what these $IPD$-specific plots don't give the reader is a sense of how $IPD$ affects within-fragment co-methylation.

Instead, for each sample I plot selected quantiles --- $10\%, 25\%, 50\%, 75\% and 90\%$ --- of each $IPD$-specific $LOR$ distribution as a function of $IPD$. I refer to these as _quantile plots_ of $LOR$ as a function of $IPD$. These quantile plots may also be stratified by additional annotations.

Key points:

- LOR decreases with IPD
- For a given IPD, a wide variation in LOR
- LOR does not go to zero over observable distance
- Variation is fairly constant across IPD
- Noise at the end is due to limited data
- CGIs are more co-methylated than non-CGI
- SE data can't cut the mustard
- Intriguing mode/plateua around 150-200 bp (NIL >= 0)
  - SE data suggests some is technical, e.g., incomplete removal of M-bias
  - Also, IMR90 PE data doesn't have it
- LOR vs. Pearson
- NIL = 0 vs. NIL >= 0
- iPSC vs. somatic

An example quantile plot for the __WHICH SAMPLE__ is shown in __FIGURE__. The vast majority of $LOR$s are positive, meaning that most 2-tuples are co-methylated. The strength of co-methylation decays roughly as $IPD$ increases, which can be seen by the downward shift in the quantiles as a function of $IPD$. I

__FIGURES__ show that the two DNA strands have near identical $LOR$ distributions. This is not to discount the possibility of 2-tuples with strand-specific differences in co-methylation, however, it does mean that it is generally safe to aggregate counts across strands.

__FIGURES__

__TODO__

\section{Summary}

- Discuss 1D nature of $IPD$s vs. $4D$ structure of genome (3D + time)

__OLD WRITING BELOW THIS LINE__

### Visualisations

>>>>

\FloatBarrier
\section{Correlations of $\beta$-values}\label{subsec:beta_cor}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/feature_status.pdf}
\caption{Schematic illustrating the 'feature status' of four pairs of methylation loci. The circles represent loci and the grey bars a genomic feature, such as CpG islands. Each pair's 'feature status' is determined by whether both loci are outside of the feature (outside), one locus is inside and one outside of the feature (spanning), or both inside the feature (inside). Note that the rightmost pair is 'inside' even though the two loci that comprise the pair are in different elements of the genomic feature.}
\label{fig:feature_status}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/ADS_beta_scatterplots.pdf}
\caption{Kernel density smoothed scatterplots for pairs of CpG $\beta$-values, $(\beta_{1}, \beta_{2})$, from the ADS sample. $\beta$-values are computed using strand-collapsed counts. $n$ is the number of pairs in each scatterplot.  The Spearman correlation of each scatterplot is also reported. Individual plots created using the \texttt{smoothScatter} function in R.}
\label{fig:ADS_beta_scatterplots}
\end{figure}

\FloatBarrier
\subsection{Results}

@@

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_5}, \ref{fig:Lister_spearman_beta_cors_min_cov_5}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_5} and \ref{fig:Ziller_spearman_beta_cors_min_cov_5} show the Spearman correlations of $\beta$-values for $NIL = 0$ pairs of CpGs stratified by strand. Since the correlations are very similar for the two strands, I decided to use the strand-collapsed $\beta$-values in all that follows.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_5}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_5.pdf}
\caption{Correlations of $\beta$-values for strand-specific pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_5}
\end{figure}

\FloatBarrier
\subsubsection{Choice of correlation coefficient}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed.pdf}
\caption{Pearson correlation versus Spearman correlations of $\beta$-values for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_pearson_vs_spearman_beta_cors_min_cov_10_strand_collapsed}
\end{figure}

\FloatBarrier
\subsubsection{$NIL = 0$}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values as a function of $IPD$ for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL = 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}
\end{figure}

\FloatBarrier
\subsubsection{$NIL \geq 0$}\label{sec:nil_geq_0_beta_cors}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI.pdf}
\caption{Spearman correlation of $\beta$-values for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CI}
\end{figure}

Comparing CpG islands to non-islands, Figures \ref{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}, we again see that the methylation levels of pairs of CpGs within islands are consistently more correlated than those outside of islands[^hepg2].

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the EPISCOPE dataset.}
\label{fig:EPISCOPE_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Lister dataset.}
\label{fig:Lister_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Seisenberger dataset.}
\label{fig:Seisenberger_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI.pdf}
\caption{Spearman correlations of $\beta$-values as a function of $IPD$ stratified by CpG island status for strand-collapsed pairs of CpGs with $NIL \geq 0$ for samples from the Ziller dataset.}
\label{fig:Ziller_spearman_beta_cors_all_min_cov_10_strand_collapsed_CGI}
\end{figure}

The $150-200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands. This may be due to  different regulation of DNA methylation outside of CpG islands or simply because CpG islands have such consistently high correlations that these periodicities are not evident.

\FloatBarrier
\section{Within-fragment co-methylation}\label{subsec:wf_cometh}

@@

Here I describe a simple method based on analysing methylation patterns at 2-tuples. For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type. This method is implemented in the `cometh` function that is part of the `MethylationTuples` software.

@@

\FloatBarrier
\subsubsection{Choice of dependence measure}\label{sec:choice_of_dependence_measure}

As discussed, I use the logarithm of the odds ratio to summarise within-fragment co-methylation. Here I explain my reasons for choosing this measure of association.

@@

There are several alternative measures of association for a $2 \times 2$ table of binary variables. An interesting review of these is given by \citet{Kraemer:2006te}. The Pearson correlation coefficient, also known as the phi coefficient when applied to binary $2 \times 2$ tables, was used by \citet{Landan:2012kp} to analyse within-fragment co-methylation. There are a few reasons that I prefer $LOR$ to Pearson's correlation.

Firstly, the ``possible range of the [log] odds ratio is unaffected by the marginal distributions'' (__CITE: quote from \cite{Kraemer:2004el} but references two other papers. Find original quote__). Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints (__CITE: Plackett?__). In contrast, Pearson's correlation of a $2 \times 2$ is restricted by the marginal probabilities (__CITE: Kraemer HC. A measure of 2 × 2 association with stable variance and approximately normal small sample distribution: planning cost-effective studies.__), which I find undesirable.

Secondly, the empirical distribution of $LOR$ for all 2-tuples in a sample  (Figure \ref{fig:EPISCOPE_lor_histogram}) is much nicer than that for Pearson's correlation coefficient (Figure \ref{fig:EPISCOPE_pearson_histogram}). Specifically, the distributions of $LOR$ for each sample are approximately unimodal and symmetric whereas the distributions of Pearson's correlation are multimodal, asymmetric, and clearly constrained by the underlying marginal probabilities. It is far easier to work with symmetric, unimodal distributions, such as when wanting to summarise these by measures of central tendency and spread, which is important in what follows.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_histogram.pdf}
\caption{Histograms of $LOR$ for 2-tuples from the EPISCOPE data (binwidth $= 0.5$).}
\label{fig:EPISCOPE_lor_histogram}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_pearson_histogram.pdf}
\caption{Histograms of Pearson correlation coefficient for 2-tuples from the EPISCOPE data (binwidth $= 0.05$).}
\label{fig:EPISCOPE_lor_histogram}
\end{figure}

Finally, and related to the second point, there are $2 \times 2$ tables that are clear evidence of co-methylation that have a Pearson correlation close to zero. An example is shown in Table \ref{tab:2x2_big_lor_small_cor}. This table has a $LOR = 3.3$ and Pearson's correlation of $-0.025$, which is due to the table being dominated by one cell's count, $uu = 39$. This occurs for all tables where one cell dominates. The problem is only partially fixed by adding an offset. For example, adding $1$ to all cells gives a Pearson's correlation of $0.29$, which is still a low correlation.

\begin{table}[h]
\centering
\begin{tabular}{lcccl}
& \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
&                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5}
\multirow{2}{*}{First CpG} & \textbf{M}           & 0              & 1             & 1            \\
& \textbf{U}           & 1              & 39             & 40             \\ \cline{2-5}
& \textbf{Total}       & 1              & 40             & 41
\end{tabular}
\caption{A $2 \times 2$ contingency table with a large $LOR$ but small Pearson's correlation.}
\label{tab:2x2_big_lor_small_cor}
\end{table}
